Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-14 paper.tex
refactored sec 3

\definecolor{lightgrey}{HTML}{FFFFFF}

@@

\section{Scalable Output Approximation}%
The main objective of this experience report is to investigate whether we can
have built practical tools that are on output approximations like the tools in
Section \ref{sec:existing_tools}. In particular, we want to find out whether
symbolic execution as the underlying technique for static output approximation
is scalable, i.e., it computes an accurate approximation for large and modern
PHP systems with reasonable time and space consumption.

@@

\subsection{Towards Scalable Output Approximation?}
As our goal is to have practical tools support, we approach this question by
re-implementing the symbolic execution semantics described in
\cite{Nguyen:2014:BCG:2635868.2635928} and extending it with support for
additional language features. We choose to aim to support symbolic execution
for a real-world example system of reasonable size that incorporates the
missing language features described in Section \ref{sec:limitations} and is
still developed and maintained: \textsf{WordPress} is a popular open-source
Content Management System (CMS) with around 300k lines that provides a vivid
plug-in environment and is maintained by a large community.
For our implementation of a symbolic execution engine for PHP,
\textsf{Oak}\footnote{The implementation can be found
at \url{github.com/smba/oak}.}, we have chosen a test-driven approach with
continuous integration using regression tests based on \textsf{WordPress} and \textsf{SchoolMate} (see Section \ref{sec:experiment_setup}). Oak incorporates the existing execution semantics
of \cite{Nguyen:2014:BCG:2635868.2635928} and extends it with support for
object-oriented orientation (class definition, class instantiation and method
invocation) as well as PHP array operations. A list of the functions provided
by the PHP standard library can be found at \cite{PHP}.
Throughout the implementation we did not achieve high code coverage for
WordPress, which was the main motivation to further investigate limitations in
the wild.

\subsection{Experience Report}
We identified a number of trade-offs between accuracy and performance that may
cause the approximation to be impractical, yet possible. Also, we identified
limitations impeding symbolic execution conceptually. For certain language
features, concrete information rather than symbolic values may be required to
actually be executed properly as this defective execution can impact subsequent
parts of the program which may become inaccessible and may be missed.

\subsubsection{Symbolic vs Concrete Execution}
Throughout the implementation, we basically followed the symbolic execution
semantics described in \cite{Nguyen:2014:BCG:2635868.2635928}. Nevertheless, for additional language features as
well as standard library functions we experienced it to be beneficial to the
approximation to execute them as concretely as possible.
For instance, \sf{WordPress} uses a text filter mechanism, where different
filter functions are applied sequentially to a string literal. All filters can
be linked to a hook using  associative arrays so that when the hook is
triggered, all filters linked to it are applied. As described in Section \ref{}
we execute loops, in particular while loops, only once due to an possibly
indeterminable number of iterations. PHP yet offers more loop constructs: the
do-while loop and the foreach loop. While for the do-while loop as well as for
the while loop the number of iterations is possibly indeterminable, the foreach
loop iterates over all items of a given array whose number of items is known at
runtime. For \sf{WordPress}, we are able to achieve a more accurate
approximation and higher code coverage results by concretely executing foreach
loops as more server-side code (filter functions in our case) are accessed and
executed. In turn, exhaustively executing foreach loops and filter functions
concretely is a trade-off between accuracy and performance, as single hook can
be called for different contexts as we are exploring multiple execution paths.

An additional aspect on the boundary of concrete and symbolic execution, or
effort and accuracy respectively, is the extent to which program inputs are
assumed to be symbolic. As shown in Section \ref{} introducing symbolic values
can be problematic if they are propagated to language features requiring
concrete information, such as include expressions or indirect function calls.
We manually scaled the number of sources of symbolic inputs by providing static
dummy inputs such as a concrete form input, file name or configuration entry.
For specific cases, where concrete information is inevitably required, we can
achieve higher code coverage and a more accurate approximation, as more code
becomes accessible. Nevertheless, while detecting dynamic features can be
automated, it requires manual inspection and understanding of the web
application to provide appropriate dummy input data. Again, this is a trade-off
between the effort to provide input data and accuracy of the approximation and
code coverage respectively.

\subsubsection{Multi-Target Method Calls}
The first non-conceptual problem we encountered, in particular for
\sf{WordPress} is related to object-oriented programming. While exploring
different execution paths when the control flow is split, a variable can have
different values depending on the path condition under which a value has been
assigned to it (see the code example in Figure \ref{fig:output_example}). Similar, a variable
can point to different objects depending on the path condition as the snippet in
Figure \ref{fig:multi_target} illustrates. The variable \tt{\$person} can either
point to object \tt{Alice} or \tt{Bob}. In addition, for any method invocated on
that variable, the actual target is ambiguous as there are two, depending on the
path condition. In order to have a sound symbolic execution, both method calls
need to be executed. Nevertheless, the number of contexts for a method can grow
rapidly, and easily become infeasible to execute.

We implemented several heuristic approaches to tame this state space explosion
for multi-target method calls, including executing a method only for a subset or
just a single object. This trade-off between accuracy and soundness
of the symbolic execution and performance time required.

\subsubsection{Dynamic Language Features}
Conceptually, symbolic execution failed for two types of language features. As
stated in Section \ref{sec:workflow}, dynamic web applications are staged and client page
output is computed at runtime. Symbolic execution hand has shown to be a useful
technique to approximate dynamically generated output, it is challenged by
language features that require concrete information rather than symbolic
information at runtime. In particular, this applies to language features that
provide access to reusable code of the web application: include expressions and
function calls.

First, include expressions in PHP usually require a string value representing
the path to the script file or template to include. If this string value is
statically provided by a string literal or can be constructed unambiguously, the
symbolic execution engine was able to resolve the include expression and include
the desired file. However, it is also common to assemble include expressions not
only static string literals, but from several sources including user input
information, database query results or configuration files. Figure
\ref{fig:dynamic_includes} shows a small example of a dynamic include containing
information retrieved from a database. Since any input for a symbolically
executed application is symbolic this information leads to include expressions
containing symbolic values. Also, configuration files represent only one
particular configuration or even just default values or placeholders. Hence, 
symbolic execuion is prone to fail for dynamic includes if symbolic information
is contained.

Second, function calls can either be direct by statically providing the name of
the function (or method), or indirect. Indirect function calls in PHP are
enabled by using built-in functions like \tt{call\_user\_func},
\tt{call\_user\_func\_array} or simply passing a string value with a direct
function call to the \tt{eval} function. The first two functions take as
arguments the name of the function two call, and all arguments to that functions either as additional arguments or
array respectively. The eval function evaluates a given PHP expression passed
as a string literal.
Again, we encountered similar issues to dynamic include expressions with
indirect function call mechanisms. The example in Figure
\ref{fig:indirect_calls} illustrates two functions \tt{header\_serif} and
\tt{header\_sserif} which both print a HTML headline. Depending on whether the
value of the variable \tt{\$style} retrieved from a database is \tt{serif} or
\tt{sserif}, a different function is called. Given that the function name is 
concrete at runtime, the desired function can be called. Otherwise, if the
function name is assembled at runtime and contains symbolic values, no function
can be called as the name is ambiguous or unknown. For this example, symbolic
execution is not able to determine which function to call indirectly.

@@

\subsubsection{Interpreter Customization}
We have approached issues we encountered in different ways, including
multiple-target method calls and foreach loops, as these represent a trade-off
between accuracy of execution and performance.
For method calls having ambiguous targets, we implemented different modes of
execution, ranging from a complete mode where a method is executed for all
possible targets to abstract modes with either a subset of targets or only a
single target. In the latter cases where targets are discarded, we assumed
symbolic values as method return values. Our symbolic interpreter can be
customized to work in the desired mode to scale accuracy or performance.
Nevertheless, for \sf{WordPress} these variations did not have significant
impact on the output approximation as symbolic execution was rather impeded by
dynamic features.
Also, for foreach loops we are able to toggle between a single iteration mode
and a complete iteration mode. Although we were able to increase code coverage
by accessing more code, it did not significantly increase our output
approximation.
Note that these variants have only been tested for \sf{WordPress} to better
understand possible trade-offs or simply achieve any output approximation in a
reasonable period of time. We believe based on our experience dynamic features
to be the main conceptual limitation impeding code coverage as well as output
approximation quality. In the next Section we analyze in detail the impact of
dynamic features with concrete foreach loop execution and complete
multiple-target execution for method calls.

>>>>

\definecolor{lightgrey}{HTML}{FFFFFC}

@@

\section{Scalable Output Approximation}%
The main objective of this experience report is to investigate whether we can
have a scalable symbolic execution engine as it is the underlying technique for
static output approximation in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}.
Our vision is a practical symbolic execution, i.e., it computes an accurate approximation for large and modern PHP systems with reasonable time and space consumption.

@@

\subsection{Towards Scalable Output Approximation?}
As our goal is to have a scalable symbolic execution engine, we approach this
question by re-implementing the symbolic execution semantics described in
previous work \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with
support for additional language features. We aim to support symbolic execution
for a real-world example system of reasonable size that incorporates the
missing language features described in Section \ref{sec:limitations} and is
still developed and maintained. \textsf{WordPress} is a popular open-source
Content Management System (CMS) with around \textcolor{blue}{300.000} lines of
code that provides a vivid plug-in environment and is maintained by a large
community.
For our implementation of a symbolic execution engine for PHP,
\textsf{Oak}\footnote{The implementation can be found
at \url{github.com/smba/oak}.}, we have chosen a test-driven approach with
continuous integration using regression tests based on \textsf{WordPress} and \textsf{SchoolMate} (see Section \ref{sec:experiment_setup}). Oak incorporates the existing execution semantics
of \cite{Nguyen:2014:BCG:2635868.2635928} and extends it with support for
object-oriented orientation (class definition, class instantiation and method
invocation) as well as most PHP array operations. A list of the functions
provided by the PHP standard library can be found at \cite{PHP}.
Throughout the implementation we did not achieve high code coverage for
\sf{WordPress}. In the following we present our observations we made thorughout
the implementation regarding issues impeding symbolic execution.

\subsection{Experience Report}
We identified trade-offs between accuracy and performance that may
cause the approximation to be impractical, yet possible. Additionally, we
detected limitations impeding symbolic execution conceptually: For certain
language features, concrete information rather than symbolic values may be required to
actually be executed properly as this defective execution can impact subsequent
parts of the program which may become inaccessible and may be missed.

\subsubsection{Symbolic vs Concrete Execution}
Throughout the implementation, we basically followed the symbolic execution
semantics described in previous work \cite{Nguyen:2014:BCG:2635868.2635928}.
Nevertheless, for additional language features we experienced it to be beneficial to the
approximation to execute them as precisely as possible.
For instance, \sf{WordPress} uses a text filter mechanism, where different
filter functions are applied sequentially to a string literal. Filters can
be linked to a hook using  associative arrays so that when the hook is
triggered, all filters that are linked to it are applied. As described in
Section \ref{} we execute loops only once due to an possibly
indeterminable number of iterations. PHP yet offers more loop constructs: the
do-while loop and the foreach loop. While for the do-while loop as well as for
the while loop the number of iterations is possibly indeterminable, the foreach
loop iterates over all items of a given array whose number of items is known at
runtime. For \sf{WordPress}, we are able to achieve a more accurate
approximation and higher code coverage results by concretely executing foreach
loops as more server-side code (filter functions in our case) are accessed and
executed. In turn, exhaustively executing foreach loops and filter functions
concretely is a trade-off between accuracy and performance, as a single hook
call site can be called from various contexts since we are exploring multiple
execution paths.

An additional aspect on the trade-off between concrete and symbolic execution,
or effort and accuracy respectively, is the extent to which program inputs are
assumed to be symbolic. As shown in Section \ref{} introducing symbolic values
can be problematic if they are propagated to language features requiring
concrete information, such as include expressions or indirect function calls.
We manually scaled the number of sources of symbolic inputs by providing static
dummy inputs such as a concrete form input, file name or configuration entry.
For specific cases, where concrete information is inevitably required, we can
achieve higher code coverage and a more accurate approximation, as more code
becomes accessible. Nevertheless, while detecting dynamic features can be
automated, it requires manual inspection and understanding of the web
application to provide appropriate dummy input data. Again, this is a trade-off
between the effort to provide input data and accuracy of the approximation and
code coverage respectively.

\subsubsection{Multi-Target Method Calls}
A non-conceptual problem we encountere is related to object-oriented programming. While exploring
different execution paths when the control flow is split, a variable can have
different values depending on the path condition under which a value has been
assigned to it (see the code example in Figure \ref{fig:output_example}).
Similarly, a variable can point to different objects depending on the path
condition as the snippet in Figure \ref{fig:multi_target} illustrates. The variable \tt{\$person} can either
point to object \tt{Alice} or \tt{Bob}. In addition, for any method invocated on
that variable the actual target is ambiguous as there are two, depending on the
path condition. In order to have a sound symbolic execution, both method calls
need to be executed. Nevertheless, the number of contexts for which a method can
be called can grow rapidly, and easily become infeasible to execute
exhaustively.

We implemented several heuristic approaches to tame this state space explosion
for multi-target method calls, including executing a method only for a subset of
target objects, or just a single object. This is a trade-off between accuracy
and soundness of the symbolic execution and performance time required.

\subsubsection{Dynamic Language Features}
Conceptually, we identified two language features for which symbolic execution
failed frequently. As stated in Section \ref{sec:workflow}, dynamic web
applications are staged and client page output is computed at runtime. Symbolic
execution has shown to be a useful technique to approximate dynamically
generated output in theory, though it is challenged by language features that
require concrete information rather than symbolic information at runtime. In particular, this
applies to language features that provide access to reusable code of the web
application: include expressions and function calls.

First, include expressions in PHP usually require a string value representing
the path to the script file or template to include. If this string value is
statically provided by a string literal or can be constructed unambiguously, the
symbolic execution engine was able to resolve the include expression and include
the desired file. However, it is also common to assemble include expressions not
only from static string literals, but from several sources including user input
information, database query results or configuration files. Figure
\ref{fig:dynamic_includes} shows a small example of a dynamic include containing
information retrieved from a database. Since any input for a symbolically
executed application is symbolic this information can be propagated to include
expressions. Also, configuration files represent only one particular
configuration or even just default values or placeholders. Hence, symbolic
execuion is prone to fail for dynamic includes if symbolic information is contained.

Second, function can either be called directly by statically providing the name
of the function (or method), or indirectly. Indirect function calls in PHP are
enabled by using built-in functions like \tt{call\_user\_func},
\tt{call\_user\_func\_array} or simply passing a string value with a direct
function call to the \tt{eval} function. The first two functions take as
arguments the name of the function two call, and all arguments to that
functions either as additional arguments or a single associative array
respectively.
The \tt{eval} function evaluates a given PHP expression passed as a string
literal.
Again, we encountered issues similar to dynamic include expressions with
indirect function call mechanisms: The example in Figure
\ref{fig:indirect_calls} illustrates two functions \tt{header\_serif} and
\tt{header\_sserif} which both print a HTML headline. Depending on whether the
value of the variable \tt{\$style} retrieved from a database is \tt{serif} or
\tt{sserif}, a different function is called. Given that the function name is 
concrete at runtime, the desired function can be called. Otherwise, if the
function name is assembled at runtime and contains symbolic values, no function
can be called as the name is ambiguous or unknown. For this example, symbolic
execution is not able to determine which function to call indirectly.

@@

\subsubsection{Interpreter Customization}
We have approached issues we encountered, including
multiple-target method calls and foreach loops, in different ways as these
represent a trade-off between accuracy of execution and performance.
For method calls having ambiguous targets, we implemented different modes of
execution, ranging from a complete mode where a method is executed for all
possible targets to abstract modes with either a subset of targets or only a
single target. In the latter cases where targets are discarded, we assumed
symbolic values as method return values. Our symbolic interpreter can be
customized to work in the desired mode to scale accuracy or performance.
Nevertheless, for \sf{WordPress} these variations did not have significant
impact on the output approximation as symbolic execution was rather impeded by
dynamic features.
Also, for foreach loops we are able to toggle between a single iteration mode
and a complete iteration mode. Although we were able to increase code coverage
by accessing more code, it did not significantly increase our output
approximation.
Note that these variants have only been tested for \sf{WordPress} to better
understand possible trade-offs or simply achieve any output approximation in a
reasonable period of time. We believe based on our experience dynamic features
to be the main conceptual limitation impeding code coverage as well as output
approximation quality. In the next Section we analyze in detail the impact of
dynamic features with concrete foreach loop execution and complete
multiple-target execution for method calls.
